Clustering Constraints on the Relative Sizes of Central and Satellite Galaxies

### 2017-11-03 paper/galsize_paper.tex
Finished draft of 5.1

Backwards modeling methods have been used extensively in the literature to gain insight into the relationship between galaxy mass, size, and environment. Broadly speaking, such studies proceed by using a galaxy group catalog to classify observed galaxies as centrals or satellites, and estimate their halo mass.  Employing such methods, several analyses of the \citet{yang_etal05b} group catalog have found that the $\mstar-\rhalf$ relation of early-type galaxies exhibits weak, if any, environmental dependence \citep{huertas_company_etal13b,shankar_etal14}.

A direct comparison to this finding is not possible because the conclusions drawn here pertain to $\mstar-$limited galaxy samples. As discussed in \citet{spindler_wake17}, it is entirely possible that size differences between centrals and satellites of the same mass can be accounted for by mutual covariance with an additional variable such as star formation rate or morphological type \citep[see also][for an explicit demonstration of this scenario]{lilly_carollo16}. In particular, suppose that centrals and satellites of the same mass have different early-type fractions as indicated in \citet{weinmann_etal06}, and that early- and late-type galaxies exhibit universal, but distinct, $\mstar-\rhalf$ relations. In such a case, centrals would be larger than satellites of the same mass, even though {\em early-type centrals} would have the same sizes as {\em early-type satellites}.

As discussed in \S\ref{subsec:future}, we are currently pursuing follow-up work in which we jointly model morphological type together with galaxy size. We consider forward modeling methods to be a requisite for progress on this issue, not only to properly handle the multi-dimensional nature of the problem, but also to rigorously treat systematic errors that plague inference based on galaxy group catalogs \citep[see][for a thorough discussion]{campbell_etal15}.

Our approach is closely aligned with the methods used in \citet{somerville_etal17}, who studied the empirical modeling features that are necessary to recover the tight scatter in the observed $\langle\rhalf\vert\mstar\rangle$ relation. By building models where $\rhalf$ is set by halo spin $\lambda_{\rm halo}$, the authors in \citet{somerville_etal17} found that the level of intrinsic scatter about $\langle\lambda_{\rm halo}\vert\mhalo\rangle$ in dark matter halos is at least as large as the scatter about $\langle\rhalf\vert\mstar\rangle$ seen in observed galaxies. Since the latter necessarily receives an additional contribution from measurement error, this implies some tension with the common semi-analytic modeling assumption that $\lambda_{\rm halo}$ scales with $\rhalf.$ As noted in \citet{somerville_etal17}, tension in the level of scatter cannot be used to directly test the $\lambda_{\rm halo}\propto\rhalf$ assumption because the physical motivation for this correlation is largely limited to disk galaxies \citep{mo_mao_white98}. This tension is not present in our approach for the reason that the level of scatter is simply a modeling parameter in our approach, and we make no attempt to uncover the physical origin of this scatter. However, our fiducial value choice was motivated by \citet{somerville_etal17}, and in the ongoing follow-up work discussed in \S\ref{subsec:future} we will systematically test the large-scale structure implications of the assumption that $\lambda_{\rm halo}\propto\rhalf^{\rm disk}.$

@@

First, the bottom right panel of Figure \ref{fig:colorclustering} indicates that the size of the bulge has only a tenuous connection to the properties of its parent dark matter halo.\footnote{Strictly speaking, the size of the bulge could in principle be closely connected to any number of dark matter halo properties, so long as those halo properties do not significantly impact two-point clustering.} Second, for disk-dominated galaxies, the observed $\rhalf-$dependent clustering suggests that at fixed $\mstar,$ the size of the disk is strongly correlated with whether or not the galaxy is a satellite. Our analysis complements \citet{weinmann_etal08}, who reach the same conclusions based on a study of galaxy group catalogs. These findings imply that bulge prominence is intimately connected with the quenching of star-formation, but that there is wide variation in star-formation histories amongst disk-dominated galaxies, in agreement with numerous previous findings \citep[see, e.g.,][]{woo_etal15,font_etal17}.

>>>>

Numerous previous analyses have employed the \citet{yang_etal05b} group catalog to ``directly" study the relationship between galaxy mass, size, and environment. For example, by treating observed groups as genuine dark matter halos, several analyses have found that the $\mstar-\rhalf$ relation of early-type galaxies exhibits weak, if any, environmental dependence \citep{weinmann_etal08,huertas_company_etal13b,shankar_etal14}. Meanwhile, \citet{weinmann_etal08} additionally find that for late-type galaxies, there is a strong correlation between disk size and central/satellite designation. Qualitatively, our clustering measurements shown in Figure \ref{fig:colorclustering} are in good agreement with these findings: once galaxies have been morphology-selected, only the clustering of disk shows significant dependence upon $\rhalf.$ Our clustering results are also consistent with the disk model presented in \citet{dutton_etal08,dutton_etal10}, which predicts that at fixed stellar mass, there is weak, positive correlation between disk size and halo mass.

As discussed in \citet{spindler_wake17}, it is entirely possible that size differences between centrals and satellites of the same mass can be accounted for by mutual covariance with an additional variable such as star formation rate or morphological type \citep[see also][for an explicit demonstration of this scenario]{lilly_carollo16}. In particular, suppose that centrals and satellites of the same mass have different early-type fractions as indicated in \citet{weinmann_etal06}, and that early- and late-type galaxies exhibit universal, but distinct, $\mstar-\rhalf$ relations. In such a case, centrals would be larger than satellites of the same mass, even though {\em early-type centrals} would have the same sizes as {\em early-type satellites}. This scenario could also explain the sign-change and weakening of $\rhalf-$dependent clustering measurements in the present work.

As discussed in \S\ref{subsec:future}, we are currently pursuing follow-up work in which we jointly model morphological type together with galaxy size. We consider forward modeling methods to be a requisite for meaningful progress on this issue, not only to properly handle the multi-dimensional nature of the problem, but also to rigorously treat systematic errors that plague inference based on galaxy group catalogs \citep[see][for a thorough discussion]{campbell_etal15}.

Our methodology is closely aligned with \citet{somerville_etal17}, who studied the empirical modeling features that are necessary to recover the tight scatter in the observed $\mean{\rhalf}{\mstar}$ relation. By building models where $\rhalf$ is set by halo spin $\lambda_{\rm halo}$, the authors in \citet{somerville_etal17} found that the level of intrinsic scatter about $\langle\lambda_{\rm halo}\vert\mhalo\rangle$ in dark matter halos is at least as large as the scatter about $\langle\rhalf\vert\mstar\rangle$ seen in observed galaxies. In our approach, the level of scatter is simply a modeling parameter whose fiducial value was motivated by \citet{somerville_etal17}. In ongoing follow-up work discussed in \S\ref{subsec:future}, we will systematically test the large-scale structure implications of the assumption that $\lambda_{\rm halo}\propto\rhalf^{\rm disk}$ using forward-modeling methods analogous to those advocated for in \citet{somerville_etal17} and employed here.

@@

First, the bottom right panel of Figure \ref{fig:colorclustering} indicates that the size of the bulge has only a tenuous connection to the properties of its parent dark matter halo.\footnote{Strictly speaking, the size of the bulge could in principle be closely connected to any number of dark matter halo properties, so long as those halo properties do not significantly impact two-point clustering.} Second, for disk-dominated galaxies, the observed $\rhalf-$dependent clustering suggests that at fixed $\mstar,$ the size of the disk is strongly correlated with whether or not the galaxy is a satellite. These findings imply that bulge prominence is intimately connected with the quenching of star-formation, but that there is wide variation in star-formation histories amongst disk-dominated galaxies, in agreement with numerous previous findings \citep[see, e.g.,][]{woo_etal15,font_etal17}.
